Synthesizing Quotient Lenses

### 2018-03-14 papers/QuotientLenses/main.tex
Bunch of comments

\section{Introduction}
Often, the same data is stored in more than one format. For example,
bibliographic data may be stored in \bibtex{} or EndNote formats,
spreadsheet data can be represented as either CSV and TSV files, and web
APIs can return data as either JSON or XML objects.  When changes are
made one version, the data in the other must be updated so the
two continue to represent the same information.
One way to synchronize such data is to use \emph{lenses}~\cite{Lenses}, 
programs that simultaneously define pairs of functions for translating
the data from the source format to the target formant (the \emph{get}
direction) and back again (the \emph{put} direction). Lenses can be
written in one of several domain-specific bidirectional languages that 
guarantee various \emph{lens laws} constraining how the \kw{put} and
\kw{get} functions must fit together.

@@

This paper applies this idea to the problem of
synthesizing {\em quotient lenses}~\cite{quotientlenses}.
Quotient lenses are lenses in which the lens laws are
loosened so that they hold modulo an equivalence relation on the source and
target data respectively; in this paper we are concerned with {\em bijective
quotient lenses} which are lenses for which Equation \ref{bijectivelenslaws}
holds modulo equivalence relations $\equiv_S$ and $\equiv_T$ defined on the source
and target data respectively:\bcp{Not too convinced about the colors.  The
  fact that lens expressions mix black and red is distracting -- my eyes
  can't easily tell which bits of a larger expression are lenses.}
\begin{equation}\label{quotientlenslaws}
\ell.\get \; (\ell.\lput \; t) \equiv_T t \text{, and } \ell.\lput \; (\ell.\get
\; s) \equiv_S s
\end{equation}
Quotient lenses are useful in situations where a programmer wishes for the
transformation defined by a lens to have the same behavior on data that differ
only in inessential details. For instance, a programmer may wish to ``quotient
away'' the number of white space characters between data items, the
capitalization of various strings, the sequence of fields in a record, or the
order of items in a list.

@@

\section{Background: Bijective String Languages}
\label{sec:background}
Before describing QREs and QRE lenses, we briefly review bijective
string lenses. Consider a bidirectional transformation that converts
between \bibtex{} citation records such as
\begin{verbatim}
   @Book {Lovelace,
   Author = "Ada Lovelace",
   Title = {Notes}
   }
\end{verbatim}
\noindent
and equivalent EndNote records like the following
\begin{verbatim}
   %0 Book
   %F Lovelace
   %A Ada Lovelace
   %T Notes
\end{verbatim}
\noindent
Boomerang's bijective string lenses are designed to define bidirectional
transformations such as this one, where the data formats can be matched in a
one-to-one manner, in this case by matching the label, author and title fields.
Boomerang encourages a compositional approach in which programmers
define simple lenses and then compose them using a variety of
combinators.

@@

let bib_to_end : bibtex $\Leftrightarrow$ end_note = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{A plain bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions (which are omitted for brevity).  
\iflastminute
\bcp{a possible readability suggestion: define a named constant for a quote
  character.  (Not sure if this makes sense, or if it improves things...)}
\fi}
\label{fig:example-lens}
\end{figure}

@@

Indeed this idea is what motivates our definition of {\em QRE lenses}.
Intuitively, our QRE lenses are bijective lenses with ``canonizers at the edges,''
just like Boomerang's quotient lenses. 
Figure~\ref{fig:attheedges} depicts the architecture of QRE lenses.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{QREs}
\caption{QRE lens from source $P$ to target $Q$.  $P$ and $Q$ are QREs, each
  consisting of a ``whole'' set ($W(P)$ and $W(Q)$) and a ``kernel'' part
  ($K(P)$ and $K(Q)$) Between the two kernels is a plain bijective lens.
  The \kw{get} function of the whole lens takes an argument from $W(P)$,
  applies the canonizer for $P$ to obtain a canonical representative in
  $K(P)$, then applies the \kw{get} of the plain lens, yielding an element
  of $K(Q)$, which is a subset of $W(Q)$.  The \kw{put} function does the
  reverse, mapping from $W(Q)$ to $K(P)$ (hence $W(P)$.  This lens is in
  normal form, with canonizers (specified by QREs) at the outer edges and an
  ordinary bijective lens in the center. \BCP{Both the figure and the
    caption are new---what do you think??}}\saz{I think that this figure is a
  big improvement.  One tiny possible tweak: color the \kw{get} and \kw{put}
  keywords in the same dark red as in the text.}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q_1$ has a type $q_1: Q_1 \Leftrightarrow Q_2$ where $Q_1$ and $Q_2$
are QREs. In the {\em get} direction, a QRE lens $q_1: Q_1 \Leftrightarrow Q_2$ uses the
source QRE $Q_1$ to compute a canonical representative for the data modulo the
equivalence relation defined by $Q_1$ and then applies the \kw{get} function of a
bijective string lens $\ell_1$ to this representative. In the {\em put} direction, $q_1$
operates similarly, but using the QRE $Q_2$ and the $\lput$ function of
$\ell_1$.

@@

\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:
\newcommand{\bsep}{\ \ \sep{} \ \ }
\begin{align*}
Q := \; & \normalize{R_1}{R_2}{f} \bsep{} \id(R) \bsep \collapse{R}{s} \bsep \squash{Q_1}{Q_2}{f} \\
 | \; & \perm{Q_1, \ldots, Q_n}{Q} \bsep Q_1 \semicolon Q_2 \bsep Q_1 \cdot Q_2 \bsep (Q_1
\sep Q_2) \bsep Q^*
\end{align*}
where $Q$ ranges over QREs, 
$R$ ranges over regular expressions, 
$f$ ranges over functions between regular languages, and 
$s$ ranges over character strings.

@@

\subsection{Bijective String Lenses}
\label{subsec:bijective-lenses}
We define the set of \textit{bijective string lenses} to be the set of
bijections between regular languages constructed using the following combinators.

@@

\subsection{Syntax of QRE Lenses}
\label{subsec:qre-lenses-syntax}
Having given a brief overview of the class of bijective string lenses, we now
introduce the class of QRE lenses. The language of QRE lenses is given by following
grammar:
$$ q := \lift(\ell) \sep  q_1 \cdot q_2 \sep \swap(q_1, q_2) \sep (q_1 \sep q_2)
\sep q^* \sep q_1 \; ; \; q_2 \sep \lquot(q, \ell) \sep \rquot(\ell, q),$$ where
$\ell$ ranges over bijective lenses.

@@

By the induction hypothesis, there exist bijective lenses
$\ell_1 : K(Q_1) \Leftrightarrow K(Q_2)$ and $\ell_2 : K(Q_3) \Leftrightarrow
K(Q_4)$ such that
\begin{center}

@@

Returning to the \bibtex{} to EndNote transformation of
Section~\ref{sec:examplesynth}, the QREs \cd{bibtex} and \cd{endnote}
describe a \bibtex{} record and an Endnote record respectively. We also had
the example pair \cd{(bib_example , end_example)} where

@@

Boomerang's canonizers allow one to canonize a regular language $R$ to by
mapping it to another regular language $S$ which may not be contained in $S$.
Formally, given sets $C$ and $B$ and an equivalence relation on $B$, Foster et
al defined a {\em canonizer} $q$ from $C$ to $B/{\equiv_B}$ to be a pair of
functions $q.\kw{canonize} : C \longrightarrow B$ and $q.\kw{choose} : B
\longrightarrow C$ such that for every $b \in B$,
$$q.\kw{canonize} \; (q.\kw{choose} \; b) \equiv_B \; b$$
This definition gives allows much more latitude for defining canonizers than
QREs. For example, if $\equiv_B$ is equal to $\kw{Tot}(B)$, the equivalence
relation that relates every element in $B$ to every other element in $B$, then
every function from $C$ to $B$ is a canonizer.

>>>>

\section{Introduction}
Often, the same data is stored in more than one format. For example,
bibliographic data may be stored in \bibtex{} or EndNote formats,
spreadsheet data can be represented as either CSV and TSV files, and web
APIs can return data as either JSON or XML objects.  When changes are
made one version, the data in the other must be updated so the
two continue to represent the same information.
One way to synchronize such data is to use \emph{lenses}~\cite{Lenses}, 
programs that simultaneously define pairs of functions for translating
the data from the source format to the target formant (the \emph{get}
direction) and back again (the \emph{put} direction). Lenses can be
written in one of several domain-specific bidirectional languages that 
guarantee various \emph{lens laws} constraining how the \kw{put} and
\kw{get} functions must fit together. \bcp{See my email about deemphasizing
  ``storing'' multiple formats.}

@@

This paper applies this idea to the problem of
synthesizing {\em quotient lenses}~\cite{quotientlenses}.
Quotient lenses are lenses in which the lens laws are
loosened so that they hold modulo an equivalence relation on the source and
target data respectively; in this paper we are concerned with {\em bijective
quotient lenses} which are lenses for which Equation \ref{bijectivelenslaws}
holds modulo equivalence relations $\equiv_S$ and $\equiv_T$ defined on the source
and target data respectively:\bcp{Not too convinced about the colors.  The
  fact that lens expressions mix black and red is distracting -- my eyes
  can't easily tell which bits of a larger expression are lenses.  I would
  suggest either:
  \begin{itemize}
  \item get and put italicized, plus
  \item either and lens expressions (completely) in some subtle color (maybe
  brown), or simply black.
  \end{itemize}
}
\begin{equation}\label{quotientlenslaws}
\ell.\get \; (\ell.\lput \; t) \equiv_T t \text{, and } \ell.\lput \; (\ell.\get
\; s) \equiv_S s
\end{equation}
Quotient lenses are useful in situations where a programmer wishes for the
transformation defined by a lens to have the same behavior on data that differ
only in inessential details. For instance, a programmer may wish to ``quotient
away'' the number of white space characters between data items, the
capitalization of various strings, the sequence of fields in a record, or the
order of items in a list.

@@

\section{Background: Bijective String Languages}
\label{sec:background}
Before describing QREs and QRE lenses, we briefly review bijective
string lenses. Consider a bidirectional transformation that converts
between \bibtex{} citation records such as
\begin{verbatim}
   @Book {Lovelace,
   Author = "Ada Lovelace",
   Title = {Notes}
   }
\end{verbatim}
\noindent
and equivalent EndNote records like the following\bcp{bad punctuation: needs
a colon}
\begin{verbatim}
   %0 Book
   %F Lovelace
   %A Ada Lovelace
   %T Notes
\end{verbatim}
\noindent
Boomerang's bijective string lenses are designed to define bidirectional
transformations such as this one, where the data formats can be matched in a
one-to-one manner, in this case by matching the label, author and title fields.
Boomerang encourages a compositional approach in which programmers
define simple lenses and then compose them using a variety of
combinators.

@@

let bib_to_end : bibtex $\Leftrightarrow$ end_note = preamble . label . author_lens . title_lens
\end{lstlisting}
\caption{A plain bijective lens \cd{bib_to_end} between data
matching \cd{bibtex} and \cd{end_note} regular 
expressions (which are omitted for brevity).  
}
\label{fig:example-lens}
\end{figure}

@@

Indeed this idea is what motivates our definition of {\em QRE lenses}.
Intuitively, our QRE lenses are bijective lenses with ``canonizers at the edges,''
just like Boomerang's quotient lenses. 
Figure~\ref{fig:attheedges} depicts the architecture of QRE lenses.
\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{QREs}
\caption{QRE lens from source $P$ to target $Q$.  $P$ and $Q$ are QREs, each
  consisting of a ``whole'' set ($W(P)$ and $W(Q)$) and a ``kernel'' part
  ($K(P)$ and $K(Q)$) Between the two kernels is a plain bijective lens.
  The \kw{get} function of the whole lens takes an argument from $W(P)$,
  applies the canonizer for $P$ to obtain a canonical representative in
  $K(P)$, then applies the \kw{get} of the plain lens, yielding an element
  of $K(Q)$, which is a subset of $W(Q)$.  The \kw{put} function does the
  reverse, mapping from $W(Q)$ to $K(P)$ (hence $W(P)$.  This lens is in
  normal form, with canonizers (specified by QREs) at the outer edges and an
  ordinary bijective lens in the center. \BCP{Both the figure and the
    caption are new---what do you think??}}\saz{I think that this figure is a
  big improvement.  One tiny possible tweak: color the \kw{get} and \kw{put}
  keywords in the same dark red as in the text.}\bcp{Can do, if we keep the
  color in the text :-).  How are we going to decide about that?}
\label{fig:attheedges}
\end{figure}
Every QRE lens $q_1$ has a type $q_1: Q_1 \Leftrightarrow Q_2$ where $Q_1$ and $Q_2$
are QREs. In the {\em get} direction, a QRE lens $q_1: Q_1 \Leftrightarrow Q_2$ uses the
source QRE $Q_1$ to compute a canonical representative for the data modulo the
equivalence relation defined by $Q_1$ and then applies the \kw{get} function of a
bijective string lens $\ell_1$ to this representative. In the {\em put} direction, $q_1$
operates similarly, but using the QRE $Q_2$ and the $\lput$ function of
$\ell_1$.

@@

\subsection{Syntax and Semantics of QREs}
Formally, the language of Quotient Regular Expressions (QREs) is given by the
following grammar:\bcp{bad punctuation}
\newcommand{\bsep}{\ \ \sep{} \ \ }
\begin{align*}
Q := \; & \normalize{R_1}{R_2}{f} \bsep{} \id(R) \bsep \collapse{R}{s} \bsep \squash{Q_1}{Q_2}{f} \\
 | \; & \perm{Q_1, \ldots, Q_n}{Q} \bsep Q_1 \semicolon Q_2 \bsep Q_1 \cdot Q_2 \bsep (Q_1
\sep Q_2) \bsep Q^*
\end{align*}
where $Q$ ranges over QREs, 
$R$ ranges over regular expressions, 
$f$ ranges over functions between regular languages, and 
$s$ ranges over character strings.

@@

\subsection{Bijective String Lenses}
\label{subsec:bijective-lenses}
We define the set of \textit{bijective string lenses} to be the set of
bijections between regular languages constructed using the following
combinators.\bcp{bad punctuation}

@@

\subsection{Syntax of QRE Lenses}
\label{subsec:qre-lenses-syntax}
Having given a brief overview of the class of bijective string lenses, we now
introduce the class of QRE lenses. The language of QRE lenses is given by following
grammar:\bcp{bad punctuation}
$$ q := \lift(\ell) \sep  q_1 \cdot q_2 \sep \swap(q_1, q_2) \sep (q_1 \sep q_2)
\sep q^* \sep q_1 \; ; \; q_2 \sep \lquot(q, \ell) \sep \rquot(\ell, q),$$ where
$\ell$ ranges over bijective lenses.

@@

By the induction hypothesis, there exist bijective lenses\bcp{every display
  for the rest of the section needs punctuation help}
$\ell_1 : K(Q_1) \Leftrightarrow K(Q_2)$ and $\ell_2 : K(Q_3) \Leftrightarrow
K(Q_4)$ such that
\begin{center}

@@

Returning to the \bibtex{} to EndNote transformation of
Section~\ref{sec:examplesynth}, the QREs \cd{bibtex} and \cd{endnote}
describe a \bibtex{} record and an Endnote record respectively. We also had
the example pair \cd{(bib_example , end_example)} where\bcp{punctuation}

@@

Boomerang's canonizers allow one to canonize a regular language $R$ to by
mapping it to another regular language $S$ which may not be contained in $S$.
Formally, given sets $C$ and $B$ and an equivalence relation on $B$, Foster et
al defined a {\em canonizer} $q$ from $C$ to $B/{\equiv_B}$ to be a pair of
functions $q.\kw{canonize} : C \longrightarrow B$ and $q.\kw{choose} : B
\longrightarrow C$ such that for every $b \in B$,\bcp{punctuation}
$$q.\kw{canonize} \; (q.\kw{choose} \; b) \equiv_B \; b$$
This definition gives allows much more latitude for defining canonizers than
QREs. For example, if $\equiv_B$ is equal to $\kw{Tot}(B)$, the equivalence
relation that relates every element in $B$ to every other element in $B$, then
every function from $C$ to $B$ is a canonizer.
